---
title: "Data607 HW5 Assignment - Tidying and Transforming Data"
author: "Ritesh Lohiya"
date: "March 3, 2018"
output: html_document
---

The chart above describes arrival delays for two airlines across five destinations. Your task is to:

(1) Create a .CSV file (or optionally, a MySQL database!) that includes all of the information above. You're encouraged to use a "wide" structure similar to how the information appears above, so that you can practice tidying and transformations as described below.

(2) Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy and transform your data.

(3) Perform analysis to compare the arrival delays for the two airlines.

(4) Your code should be in an R Markdown file, posted to rpubs.com, and should include narrative descriptions of your data cleanup work, analysis, and conclusions.

Please include in your homework submission:
??? The URL to the .Rmd file in your GitHub repository. and
??? The URL for your rpubs.com web page.

```{r}
library(knitr)
library(stringr)
library(tidyr)
library(dplyr)
library(ggplot2)
```

Reading the csv file:

```{r}
airlines <- read.csv("https://raw.githubusercontent.com/Riteshlohiya/Data607_Week5_Assignment_Tidying_Transforming_Data/master/Numbersense.csv", header=FALSE, sep=",", stringsAsFactors=FALSE)
airlines
```

Data cleaning:

Renaming the columns:

```{r}
airlines$V1[1] <- "airline"
airlines$V2[1] <- "status"
airlines
```

Making the row #1 as column name:

```{r}
names(airlines) <- airlines[1,]
airlines
```

Delete row #1 and 4:

```{r}
airlines <- airlines[-c(1,4), ]
airlines
```

Addind Airline names to the missing records:

```{r}
airlines$airline[2] = 'ALASKA'
airlines$airline[4] = 'AM WEST'
```

Transforming the table by using gather():

Gather: takes multiple columns and collapses into key-value pairs, duplicating all other columns as needed. You use gather() when you notice that you have columns that are not variables.

```{r}
airlines <-gather(airlines, city, number, 3:7, factor_key=TRUE)
airlines
```

Using spread() function to separate the data on Status:

Spread: Spread a key-value pair across multiple columns.

```{r}
airlines <-  airlines %>%
                 spread(status, `number`)
airlines
```


```{r}
for (i in 3:dim(airlines)[2]){
  airlines[,i] <- as.integer(airlines[,i])
}
airlines
```

Using mutate for adding new columns like total flights, % on time and % delays:

Mutate: Mutate adds new variables and preserves existing

```{r}
airlines <- airlines %>% 
  mutate(total = delayed + `on time`) 
airlines
```

```{r}
airlines <- airlines %>% 
  mutate(on_time_percentage = (`on time`/total)*100,
       delayed_percentage = (delayed/total)*100) %>%
  arrange(city)
airlines

```

Using ggplot to understand more:

```{r}
ggplot(airlines, aes(x = airline, y=delayed_percentage, fill = city)) +  
          geom_bar(stat="identity",position="dodge") + 
          xlab("Airlines") + ylab("Delays") 
```

To calculate the average:

```{r}
airlines_summary = airlines %>% group_by(airline) %>% summarise(total_delayed = sum(delayed), total_ontime = sum(`total`), avg_delay = total_delayed/total_ontime*100)
airlines_summary
```

Calculate the city totals:

```{r}
airlines %>%
  group_by(city) %>%
  summarize(count = n(), sum_city = sum(total))
```

Boxplot to check the outlier:

```{r}
ggplot(airlines, aes(airline, total)) + geom_boxplot()
```


 







